---
title: "Investment Calculator"
subtitle: "A First Pass at Retirment Planning"
format: 
  html: 
    code-fold: true
    code-summary: "show code"
    toc: true
---
```{r}
#| echo: false
#| message: false
library(tidyverse)
```

## A better investment calculator
I've been working on building a better investment calculator to enable myself and others to get a better idea of where they need to be in order to reach their retirement goals. To be sure, this isn't fully fleshed out so you can see different accounts or assign different rates to different pools of money. That said, what this calculator allows you to do which is distinct from other basic investment calculators like the ones offered by [Nerdwallet](https://www.nerdwallet.com/calculator/investment-calculator) or [Bankrate](https://www.bankrate.com/retirement/roi-calculator/) is that you can input different rates of return. Here, a "bear" case is an undesirable rate of return whereas a "base" case is likely and a "bull" case is optimistic. 

## What I'll be adding

### A 'Financial Goal' Input

In the FIRE community and elsewhere, you might hear about this as being **Your Number**. `Your Number` usually means the amount of money you need to retire or be financially independent. But how do you go about knowing `Your Number`? At the very least, you'll want to know or estimate four things. 

* The annual rate of inflation
* Year until you retire
* Desired annual spending
* A safe withdrawal rate


```{r}
#| label: storing values for retirement planning
dob <- as_date("1985-08-30") ## date of birth
retire_date <- dob %m+% years(60) # here we take dob and add 60 years to it in case you have a retirement age in mind and don't think about years to retirement
years_til_retirement <- round(time_length(interval(Sys.Date(), retire_date), unit = "years"), 0) # calculating years to retirement... long way but useful. 
inflation_rate <- 1.04 # projected inflation rate over that timespan. 

dollars_future <- round(inflation_rate^years_til_retirement, 2)
spend_retirement_today <- 150000
spend_retirement_future <- spend_retirement_today*dollars_future
portfolio_needed <- spend_retirement_future*25
```

Folks do this differently, but I start off with the annual rate of inflation and the number of years I have until I retire. The reason: If I know those the things, I can assign a value to what my future dollar is worth compared to now. We don't need to go all the way into it but what inflation does is the erode the purchasing power of a dollar over time. In other words, it will take more *future dollars* to replace the purchase power of *today's dollar*. Here's the formula for calculating how many future dollars you'll need to make up for one current dollar `(1+ 'inflation rate') ^ 'years to retirement' = 'future dollars needed'`. So if you assumed a `4%` rate of inflation and you have 20 years until retirement, the formula would look like `1.04^20`. If you run those numbers, you know you'll need `r dollars_future` in 20 years to make up the purchasing power of \$1 today. 

It's important to note that The Federal Reserve's target rate is `2%`. Some [crypto folks](https://x.com/APompliano/status/1958958008249074006) don't believe the Fed will keep that target rate. While I tend to think folks like this are mostly full of it, I just don't see inflation returning to that long-term target rate. And if it does? Well, that just means I'll have more money come retirement. Not a bad problem to have. 

So now that we know that you'll need more dollars in the future to have the same purchasing power today, it's worth spending some time thinking about how much you'd like to spend in a year to live the kind of life you'd like to live. And remember, it's retirement. You may have fewer expenses in the future because you won't be paying for daycare or maybe you'll have paid off your mortgage and will only be on the hook for property taxes and insurance. Regardless, think about how much money you'd want annually **today** and then multiply that by the inflation multiple we identified above assuming a `4%` interest rate with 20 years til retirement: `r dollars_future`. For my wife and I, our goal is to be able to spend approximately `r scales::label_dollar(accuracy = 0.01)(spend_retirement_today)`/year in today's dollars in retirement which means I'd want to be able to withdraw `r scales::label_dollar(accuracy = 0.01)(spend_retirement_future)` per year in retirement. 

What's left? Identifying a safe withdrawal rate. Bigger Pockets Money and ChooseFI have done a few great pockets on what a safe withdrawal rate is and how to think about it. I prefer back of the envelope math and will stick with the `4%` rule which means I can effectively withdraw 4% of my portfolio per year with a very small risk of running out of money. Caveats not withstanding, this means my portfolio would need to be 25 times the size of my desired withdrawal amount or `r scales::label_dollar(accuracy = 0.01)(portfolio_needed)`.

To be sure, that number assumes we'll have no other income. But we will. I know folks say that Social Security is going away but as for now it looks like Social Security will have *reduced* rather than no benefits. [Social Security estimates that it will have to reduce benefits by 23%, a far cry from being cut to `0`](https://www.npr.org/2025/06/18/nx-s1-5436828/social-security-benefits-cut-congress). You can also supplement your income with investment property if you're in that game or a part-time job. Regardless, I want this tool to help paint a picture of the trajectory you're on to reach your financial independece or retirement goals. 

**[Disclosure]:** I'm new to this so please bear with me as I write and rewrite this page. Maybe I'll turn it into a blog. I don't know. Enjoy!

<div style="text-align: center;">

## Investment Calculator

</div>

:::{.column-page}
```{r}
#| echo: false
#| results: asis
htmltools::tags$iframe(
  src = "standalone_invest_calc.html",
  width = "100%",
  height = "800px",
  style = "border:none;"
)
```
:::
