---
title: "Massachusetts"
author: "Ardeshir Pezeshk"
format: 
  html: 
    code-fold: true
    code-summary: "show code"
---

```{r}
#| label: packages
#| message: false
#| echo: false
library(tidyverse)
library(readxl)
library(plotly)
library(httr)
library(jsonlite)
library(crosstalk)
```

## Summary
If you spend enough time with homeowners, you're likely to hear them gripe about property taxes. And as you'd expect, the volume of griping increases with an increase in the amount of property taxes paid. 

But even if you're not a homeowner, property taxes are a big deal. If you're home shopping, along with principal, interest, and private mortgage insurance, it's one of the four biggest line items in your estimated mortgage payment. Perhaps more importantly, as a resident of whatever municipality, property taxes are often the largest contributor of revenue. 

In an effort to aid research and public conversation, I wanted to build a usable tool for folks to look at property tax rates across time and space. The interactive tables and graphics below aim to do just that. 

Data for property tax rate for Massachusetts municipalities can be found [here](https://dls-gw.dor.state.ma.us/reports/rdpage.aspx?rdreport=propertytaxinformation.taxratesbyclass.taxratesbyclass_main) and the data for municipality geometry can be found [here](https://gis.data.mass.gov/datasets/43664de869ca4b06a322c429473c65e5_0/explore?location=42.051370%2C-71.715950%2C9.42)

## But What are property taxes?

Property taxes are tied to the value of a home and are paid by the homeowner. In Massachusetts, The total *amount paid* is a function of the **assessed value** of a home and the property tax rate. It's important to note that the *assessed value* of a home is not the same as the *cost* of the home. For example, in California, property taxes are largely a function of the purchasing price and rate. But in Massachusetts, the *purchasing price* of a home and its *assessed value* are distinct with the *assessed value* often being lower than the purchasing price. 

Why the discrepancy between assessed value and purchase price? The major reason in Massachusetts is tied to the passage of Proposition 2 1/2 which passed in 1980 and went into effect in 1982. The state provides [a number of YouTube videos(!)](https://www.mass.gov/info-details/proposition-2-12-and-tax-rate-process) that explain the proposition and its effects. That said, the TLDR is that it limited the annual increase in property tax revenue municipalities could levy by 2.5%. For example, if my town of Northampton raised \$1,000,000 in property tax revenue, the max it could raise the following year would be \$1,025,000.

## Visualizing Rates Across Massachusetts in the Current Fiscal Year
The map below highlights rates across all municipalities in the state of Massachusetts. 
```{r}
#| label: tax and geo data + plotly figure
#| message: false
# path to data
tax_data_path <- "../../../Library/CloudStorage/GoogleDrive-pezeshka@gmail.com/My Drive/R/interests/data/"

# read in property tax data
tax_prop <- read_excel(paste0(tax_data_path, "taxratesbyclass.xlsx")) %>% 
  rename_with(tolower) %>% 
  rename_with(~str_replace_all(., " ", "_")) |> 
  mutate(across(c(dor_code, fiscal_year), ~as.numeric(.))) %>% 
  filter(municipality!="Devens") # not in geojson data

# read in geometry data for the municipalities
library(sf)
geojson_data <- read_sf(paste0(tax_data_path, "ma_municipalities.geojson")) %>% 
  rename_with(tolower) %>% 
  mutate(county = str_to_title(county))

# merge tax with geo data
merged_df <- tax_prop |> 
  left_join(geojson_data, 
            join_by(dor_code==town_id)) 

map_df <- merged_df %>% 
  filter(fiscal_year==2025, 
         !is.na(residential)) %>% 
  select(municipality, residential, county, geometry)

# make mappable object. 
map_sf <- st_as_sf(map_df)

fig <- 
  plot_ly(
    map_sf,
    split = ~municipality,
    color = ~residential,            # fill color by rate
    showlegend = FALSE,
    stroke = I("white"),             # Set the border color
    span = I(0.3),                   # Set the border width
    text = ~paste0(                  # Define hover text
      "<b>", municipality, "</b><br>",
      "Residential Tax Rate: $", round(residential, 2), "</b><br>",
      "County: ", county
    ),
    hoverinfo = "text",
    hoveron = "fills"
  ) %>%
  hide_colorbar() %>%
  layout(
    title = "2025 Residential Property Tax Rates by Municipality<br>Per $1,000 in Assessed Value",
    geo = list(visible = FALSE)
  )

fig
```

## Property Tax Rates Over Time
Below I've visualized Property Tax Rates over time. The filter allows you to choose one or counties as your starting pot. By default it is set to 'Hampshire' county. From there, you can select and deselect towns and compare. 
```{r}
library(crosstalk)
# need county in tax_prop 
shared_df <- SharedData$new(merged_df, key = ~county)

gg_rates <- ggplot(shared_df, aes(x = fiscal_year, y = residential, color = municipality, group = municipality)) +
  geom_line(linewidth = 1) + 
  geom_point(size = 2.5) +
  labs(
    title = "Residential Property Tax Rates Over Time",
    x = "Year",
    y = "Rate per $1,000 Assessed Value",
    color = "Municipality" # Sets the legend title
  ) +
  theme_minimal(base_size = 14)

filter_select(
  id = "county_filter",
  label = "Select County or Counties:",
  sharedData = shared_df,
  group = ~county,
  multiple = TRUE,
  all = FALSE
)

remove_buttons <- c("zoom", 
                    "pan", 
                    "select", 
                    "zoomIn", 
                    "zoomOut", 
                    "autoScale", 
                    "lasso2d", 
                    "hoverClosestCartesian")

ggplotly(gg_rates, 
         tooltip = c("fiscal_year", "y", "municipality", "county")) %>%
  layout(
    legend = list(title = list(text = '<b>Municipality</b>')), 
    annotations = list(
      x = 1, # X-coordinate (1 means right edge of the plot paper)
      y = -0.1, # Y-coordinate (-0.1 means slightly below the bottom edge of the plot paper)
      text = "Source: MA Department of Revenue", # The caption text
      showarrow = FALSE, # Hide the arrow
      xref = "paper", # X-reference to the plot paper
      yref = "paper", # Y-reference to the plot paper
      xanchor = "right", # Anchor the text to the right
      yanchor = "auto", # Auto-adjust y-anchor
      font = list(size = 10, color = "grey"))) %>% 
  plotly::config(modeBarButtonsToRemove = remove_buttons, 
                 displaylogo = FALSE)
```
```{js}
function filter_default() {
    document.getElementById("county_filter").getElementsByClassName("selectized") 
[0].selectize.setValue("Hampshire", false);
 }
window.onload = filter_default;
```
```
