---
title: "MA Property Tax"
author: "Ardeshir Pezeshk"
format:
  html: 
    code-fold: true
    code-summary: "show code"
    toc: true
---

```{r}
#| label: packages
#| message: false
#| echo: false
library(tidyverse)
library(readxl)
library(plotly)
library(httr)
library(jsonlite)
library(crosstalk)
```

## Summary
If you spend enough time with homeowners, you're likely to hear them gripe about property taxes. And as you'd expect, the volume of griping increases with an increase in the amount of property taxes paid. 

But even if you're not a homeowner, property taxes are a big deal. If you're home shopping, along with principal, interest, and private mortgage insurance, it's one of the four biggest line items in your estimated mortgage payment. Perhaps more importantly, as a resident of whatever municipality, property taxes are often the largest contributor of revenue. 

The questions of who to tax, at what rate, and for what purpose are at the core local, state, and national politics. In my own town, the local paper has been filled with news and commentary regarding funding for local schools, the poor quality of roads, and the budget more generally. It's even a flashpoint in this year's [municipal election](https://www.gazettenet.com/Guest-columnist-Simon-62626497). 

In an effort to ground public conversation in data, I wanted to build a usable tool for folks to look at property tax rates across time and space. The interactive tables and graphics below aim to do just that. 

### A note on data
Data for property tax rate for Massachusetts municipalities can be found [here](https://dls-gw.dor.state.ma.us/reports/rdpage.aspx?rdreport=propertytaxinformation.taxratesbyclass.taxratesbyclass_main) and the data for municipality geometry can be found [here](https://gis.data.mass.gov/datasets/43664de869ca4b06a322c429473c65e5_0/explore?location=42.051370%2C-71.715950%2C9.42)

## But What are property taxes? {#sec-taxes}

Property taxes are tied to the value of a home and are paid by the homeowner. In Massachusetts, The total *amount paid* is a function of the **assessed value** of a home and the property tax rate. It's important to note that the *assessed value* of a home is not the same as the *cost* of the home. For example, in California, property taxes are largely a function of the purchasing price and rate. But in Massachusetts, the *purchasing price* of a home and its *assessed value* are distinct with the *assessed value* often being lower than the purchasing price. 

Why the discrepancy between assessed value and purchase price? The major reason in Massachusetts is tied to the passage of Proposition 2 1/2 which passed in 1980 and went into effect in 1982. The state provides [a number of YouTube videos(!)](https://www.mass.gov/info-details/proposition-2-12-and-tax-rate-process) that explain the proposition and its effects. [There's also a not-so-easy to follow primer](https://www.mass.gov/doc/levy-limits-a-primer-on-proposition-2-12-0/download#:~:text=Proposition%2021%E2%81%842%20allows%20a%20community%20to%20assess%20taxes,of%202.5%20percent%20each%20year.). That said, the TLDR is that it limited the annual increase in property tax revenue municipalities could levy by 2.5%. For example, if my town of Northampton raised \$1,000,000 in property tax revenue, the max it could raise the following year would be \$1,025,000. There are exceptions for new growth and overrides but that's the jist. 

## Visualizing Rates Across Massachusetts in the Current Fiscal Year
The map below highlights rates across all municipalities in the state of Massachusetts. The lighter the color, the lower the rate.
```{r}
#| label: tax and geo data + plotly figure
#| message: false
#| warning: false
# path to data
data_path <- "../../../Library/CloudStorage/GoogleDrive-pezeshka@gmail.com/My Drive/R/interests_github/data/"

# read in property tax data with geo. Saved as R file
tax_prop <- read_rds(file = paste0(data_path, "clean_tax_geo_data.R"))

# read in geometry data for the municipalities
library(sf)
map_sf <- st_read(paste0(data_path, "clean_ma_geo.geojson"), quiet = TRUE)

fig <- 
  plot_ly(
    map_sf,
    split = ~municipality,
    color = ~residential,            # fill color by rate
    showlegend = FALSE,
    stroke = I("white"),             # Set the border color
    span = I(0.3),                   # Set the border width
    text = ~paste0(                  # Define hover text
      "<b>", municipality, "</b><br>",
      "Residential Tax Rate: $", round(residential, 2), "</b><br>",
      "County: ", county
    ),
    hoverinfo = "text",
    hoveron = "fills"
  ) %>%
  hide_colorbar() %>%
  layout(
    title = "2025 Residential Property Tax Rates by Municipality<br>Per $1,000 in Assessed Value",
    geo = list(visible = FALSE)
  )

fig
```

## Property Tax Rates Over Time
Below I've visualized Property Tax Rates over time. The filters below allow you to choose one or more municipalities or counties. You can also narrow the time range by using the slider filter for year. By default the municipality filter is set to `Northampton` with the county blank and showing all available data. 
```{r}
library(crosstalk)
# need county in tax_prop 
shared_df <- SharedData$new(tax_prop)

gg_rates <- ggplot(shared_df, aes(x = fiscal_year, y = residential, color = municipality, group = municipality)) +
  geom_line(linewidth = 1) + 
  geom_point(size = 2.5) +
  labs(
    title = "Residential Property Tax Rates Over Time",
    x = "Year",
    y = "Rate per $1,000 Assessed Value",
    color = "Municipality" # Sets the legend title
  ) +
  theme_minimal(base_size = 14)

bscols(
  widths = c(4, 4, 4), 
  filter_select(
    id = "municipality_filter", 
    label = "Select Municipality/ies", 
    sharedData = shared_df, 
    group = ~municipality, 
    multiple = TRUE, 
    all = FALSE
  ), 
  filter_select(
    id = "county_filter",
    label = "Select County or Counties:",
    sharedData = shared_df,
    group = ~county,
    multiple = TRUE,
    all = FALSE
  ), 
  filter_slider(
      id = "year_filter",
      label = "Select Year Range:",
      sharedData = shared_df,
      column = ~fiscal_year, # Filter by the 'year' column
      step = 1, # Move the slider in 1-year increments
      sep = "" # Removes the comma from thousands in year (e.g., 2,024)
    )
  )

remove_buttons <- c("zoom", 
                    "pan", 
                    "select", 
                    "zoomIn", 
                    "zoomOut", 
                    "autoScale", 
                    "lasso2d", 
                    "hoverClosestCartesian")

ggplotly(gg_rates, 
         tooltip = c("fiscal_year", "y", "municipality", "county")) %>%
  layout(
    legend = list(title = list(text = '<b>Municipality</b>')), 
    annotations = list(
      x = 1, # X-coordinate (1 means right edge of the plot paper)
      y = -0.1, # Y-coordinate (-0.1 means slightly below the bottom edge of the plot paper)
      text = "Source: MA Department of Revenue", # The caption text
      showarrow = FALSE, # Hide the arrow
      xref = "paper", # X-reference to the plot paper
      yref = "paper", # Y-reference to the plot paper
      xanchor = "right", # Anchor the text to the right
      yanchor = "auto", # Auto-adjust y-anchor
      font = list(size = 10, color = "grey"))) %>%
  plotly::config(modeBarButtonsToRemove = remove_buttons, 
                 displaylogo = FALSE)
```
```{js}
function filter_default() {
    document.getElementById("municipality_filter").getElementsByClassName("selectized") 
[0].selectize.setValue("Northampton", false);
 }
window.onload = filter_default;
```

Northampton's property tax rate peaked in 2022 at \$17.89 per \$1,000 of assessed value following the pandemic and the dramatic rise in home prices. The rate has since dropped by ~22% with the 2025 Fiscal Year rate at \$13.89. 

But just because the rate drops does not mean that the total amount paid will decline. That's because there's a dance between assessed value and the tax rate. If the assessed value significantly climbs, keeping the rate constant might violate Proposition 2.5 (see @sec-taxes), overrides and new construction notwithstanding. 

And so down the rabbit hole we go to get a better understanding of the number of properties, the assessed value of all properties, and the how the ratio of assessed value to parcel or property counts have changed over time. 

## All About Parcels

Note: Parcels are not the same as units. For example, in the parcel data the state mixes together apartments that can have as few as four units with no explicit upward bound. The same may be true for condos.[^1] 

[^1]: I will update when I hear back from the Data Analytics and Resources Bureau.

... still writing this section so check back in the next few days! Today = `r Sys.Date()`.

```{r}
#| label: read in parcel count data
#| message: false
parcel_count_df <- read_csv(paste0(data_path, "parcel_counts.csv"))
```

